Implicit Neural Representation for Potential Field Geophysics

### 2023-10-10 manuscript.tex
Update per DW feedback

This thesis reports on three bodies of work across two topics in geophysics; enhancing the high-frequency content of gridded potential field data using \emph{super-resolution}; and, learning a representative function of a potential field extent using \emph{implicit neural representation}.
The first topic enhances the value of geophysical surveys by predicting high-frequency components in sparsely sampled potential field grids.
The second presents a straightforward neural network framework for high quality grid regularisation, processing, and data integration.
For each body of work in these topics, real-world geophysical case studies are presented in order to demonstrate the effectiveness and challenges of the method with open access data.
Each body of work extends deep learning methods described in computer science literature the same year as the development of each respective geophysical method, to keep pace with rapid advances in the field of machine learning.

@@

\vfill{}
\epigraph{\emph{All collected data had come to a final end. Nothing was left to be collected.}}{}

>>>>

This thesis reports on three bodies of work across two topics in geophysics: enhancing the high-frequency content of gridded potential field data using \emph{super-resolution}; and, learning a representative function of a potential field extent using \emph{implicit neural representation}.
The first topic enhances the value of geophysical surveys by predicting high-frequency components in sparsely sampled potential field grids.
The second presents a straightforward neural network framework for high quality grid regularisation, processing, and data integration.
For each body of work in these topics, real-world geophysical case studies are presented in order to demonstrate the effectiveness and challenges of the method with open access data.
Each body of work extends deep learning methods described in computer science literature the same year as the development of each respective geophysical method, to keep pace with rapid advances in the field of machine learning.

@@

\vfill{}
\epigraph{\emph{All collected data had come to a final end. Nothing was left to be collected.} \dots}{}

### 2023-10-10 paper2.tex
Make compilable
Cleverref and subfiles seem to interact poorly, and I can't cref labels in external files in the way I had set up.

I've moved to using  \include.
Update per DW feedback

\documentclass[manuscript.tex]{subfiles}
\begin{document}
\setcounter{chapter}{2}
\title{Super-resolution of Aeromagnetic Survey Data}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
\maketitle{}

@@

Photographs and other natural images of interest to computer vision are typically created by capturing light from an image sensor or are explicitly mapped on a grid using a raster graphics program.
In these rasters, each pixel is regularly distributed, and the resolution is isotropic and a factor of the pixel density of the sampling sensor or the software defined grid.
These data are encoded to regular arrays of pixel values, commonly as three-channel rasters representing the red, green, and blue intensities of each pixel.
For natural images, a common approach to generating low-resolution data is to interpolate the high-resolution image to a smaller size using a transform such as bicubic downsampling.
Aerogeophysical grids are distinct from these image rasters due to the potential field sampling and grid interpolation process outlined in \Cref{sec:resinsurveys}, and the extent to which widely spaced low-resolution geophysical line data can be enhanced, and characteristics of the resulting super-resolved grids remain to be fully explored.
Most critically, aeromagnetic data are highly sampled in the direction of flight lines but are under-sampled in the line perpendicular direction.
The resulting grid rasters may have isotropic cell size, but the frequency content in each axis is anisotropic, being a factor of the sampling frequency in the line-perpendicular (low) and line-parallel (high) directions.
Being sampled as a single component of the vector potential field, these data contain a single channel, and the data contain multiple sources of noise components.
Furthermore, the features present in these data are dependent on the subsurface geology, varying between smooth domains without significant near-surface magnetisation, through to highly deformed igneous stratigraphic units with complex high-intensity structures.
The recorded intensity values have high dynamic range, spanning tens of thousands of nanotesla, while contextually important features may be defined by tens of nanotesla \parencite{kovesiPhasePreservingTone2012}.
Because of these differences, pre-existing super-resolution models trained on large image corpuses are not suitable for predicting higher frequency signals in geophysical grids.

@@

\begin{figure}[hbt]
    % \includegraphics[width=\textwidth]{fig/p2/image1.png}
    \caption[The LTE network]{
        A simplified representation of the super-resolution LTE network.
        The encoder is marked as \(E_\phi{}\).
        \(f_\Theta{}\) is an MLP network used as the decoder for image reconstruction.
        A long-range skip connection with simple upscaling is marked with an arrow.
        The LTE network block performs high-resolution upscaling of extracted features in the Frequency domain.
        Simplified from \textcite{leeLocalTextureEstimator2022}.
    }
    \label{fig:ltenet}
\end{figure}

@@

The \(n = 1\) case is the high-resolution target (HR) with every fourth line of GT sampled at a nominal line spacing of \qty{80}{\m}, similar to real-world regional high-resolution airborne geophysical surveys.
These sampled line data are gridded at a cell size of \qty{20}{\m} using the cubic method from Verde \parencite{uieda2018}, which is one quarter of the line spacing, following the convention of \textcite{reidAeromagneticSurveyDesign1980}.
It is important to note that while the HR grid has a cell size of \qty{20}{\m}, it is not equivalent to the \qty{20}{\m} cell size GT grid, because unsampled cells are interpolated during the gridding process.
The result is a loss of high frequency detail in the line perpendicular direction, as well as the loss of features present only in the unsampled lines.

@@

\begin{figure}[hbt]
    % \includegraphics[width=\textwidth]{fig/p2/image1.png}
    \caption[Low-resolution geophysics grids]{
        The result of the proposed sampling and gridding on a selection of synthetic tiles loaded from the collection of \parencite{jessellNoddyverseMassiveData2022} at 4x scale.
        Note the coloured lines and accompanying tick marks in the ground-truth grid (GT, left) indicate cells which are sampled and used for gridding in the low-resolution grid (LR, right).
        High-resolution (HR, middle) is sampled and gridded from GT, with survey lines indicated by minor tick marks.
        The minimum curvature gridding process results creates a variety of low-resolution degradations.
    }
    \label{fig:lrdata}
\end{figure}
%TODO

@@

\subsection{Quantitative Measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, Feature Similarity (FSIM) \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image, because it is a full reference image similarity metric that quantifies structural similarity using phase and local gradient magnitude components.
%TODO complain about FSIM

\section{Results}
\label{sec:2results}
It is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.
Qualitative results are shown to compare the outputs from the two models, the baseline model that is trained with synthetic data and the fine-tuned model trained with open access survey data.

@@

\begin{figure}[hbt]
    \includegraphics[width=\textwidth]{fig/p2/srcomp_23.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata23}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\textwidth]{fig/p2/srcomp_14.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata14}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\textwidth]{fig/p2/srcomp_17.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata17}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=\textwidth]{fig/p2/srcomp_18.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata18}
\end{figure}

>>>>

% % \documentclass[manuscript.tex]{subfiles}
% \documentclass[12pt,a4paper]{report} %,openright,twoside
% \usepackage{thesisstyle}
% \addbibresource{bib/PhD.bib}
% % \setcounter{chapter}{2}
% \begin{document}

\title{Super-resolution of Aeromagnetic Survey Data}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\author[2]{Naveed Akhtar}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
% \maketitle{}

@@

Photographs and other natural images of interest to computer vision are typically created by capturing light from an image sensor or are explicitly mapped on a grid using a raster graphics program.
In these rasters, each pixel is regularly distributed, and the resolution is isotropic and a factor of the pixel density of the sampling sensor or the software defined grid.
These data are encoded to regular arrays of pixel values, commonly as three-channel rasters representing the red, green, and blue intensities of each pixel.
For natural images, a common approach to generating low-resolution data is to interpolate the high-resolution image to a smaller size using a transform such as bicubic downsampling.
Aerogeophysical grids are distinct from these image rasters due to the potential field sampling and grid interpolation process outlined in \Cref{sec:resinsurveys}, and the extent to which widely spaced low-resolution geophysical line data can be enhanced, and characteristics of the resulting super-resolved grids remain to be fully explored.
Most critically, aeromagnetic data are highly sampled in the direction of flight lines but are under-sampled in the line-perpendicular direction.
The resulting grid rasters may have isotropic cell size, but the frequency content in each axis is anisotropic, being a factor of the sampling frequency in the line-perpendicular (low) and line-parallel (high) directions.
Being sampled as a single component of the vector potential field, these data contain a single channel, and the data contain multiple sources of noise components.
Furthermore, the features present in these data are dependent on the subsurface geology, varying between smooth domains without significant near-surface magnetisation, through to highly deformed igneous stratigraphic units with complex high-intensity structures.
The recorded intensity values have high dynamic range, spanning tens of thousands of nanotesla, while contextually important features may be defined by tens of nanotesla \parencite{kovesiPhasePreservingTone2012}.
Because of these differences, pre-existing super-resolution models trained on large image corpuses are not suitable for predicting higher frequency signals in geophysical grids.

@@

\begin{figure}[hbt]
    \includegraphics[width=\linewidth]{fig/p2/temp_lte.png}
    \caption[The LTE network]{
        A simplified representation of the super-resolution LTE network.
        The encoder is marked as \(E_\phi{}\).
        \(f_\Theta{}\) is an MLP network used as the decoder for image reconstruction.
        A long-range skip connection with simple upscaling is marked with an arrow.
        The LTE network block performs high-resolution upscaling of extracted features in the Frequency domain.
        Simplified from \textcite{leeLocalTextureEstimator2022}.
    }
    \label{fig:ltenet}
\end{figure}

@@

The \(n = 1\) case is the high-resolution target (HR) with every fourth line of GT sampled at a nominal line spacing of \qty{80}{\m}, similar to real-world regional high-resolution airborne geophysical surveys.
These sampled line data are gridded at a cell size of \qty{20}{\m} using the cubic method from Verde \parencite{uieda2018}, which is one quarter of the line spacing, following the convention of \textcite{reidAeromagneticSurveyDesign1980}.
It is important to note that while the HR grid has a cell size of \qty{20}{\m}, it is not equivalent to the \qty{20}{\m} cell size GT grid, because unsampled cells are interpolated during the gridding process.
The result is a loss of high frequency detail in the line-perpendicular direction, as well as the loss of features present only in the unsampled lines.

@@

\begin{figure}[hbt!]
    \begin{center}
        \includegraphics[width=0.75\linewidth]{fig/p2/temp_lr1.png}
        \includegraphics[width=0.75\linewidth]{fig/p2/temp_lr2.png}
    \end{center}
    \caption[Low-resolution geophysics grids]{
        The result of the proposed sampling and gridding on a selection of synthetic tiles loaded from the collection of \parencite{jessellNoddyverseMassiveData2022} at 4x scale.
        Note the coloured lines and accompanying tick marks in the ground-truth grid (GT, left) indicate cells which are sampled and used for gridding in the low-resolution grid (LR, right).
        High-resolution (HR, middle) is sampled and gridded from GT, with survey lines indicated by minor tick marks.
        The minimum curvature gridding process results creates a variety of low-resolution degradations.
    }
    \label{fig:lrdata}
\end{figure}
%TODO

@@

\subsection{Quantitative Measure}
Due to the dynamic range of magnetic anomaly grids and the relative importance of structural accuracy for gridded data, image quality assessment such as peak signal-to-noise ratio or mean-squared error are an inadequate quantitative measure of upscaling performance for geophysics.
In addition to mean-square error, FSIM \parencite{linzhangFSIMFeatureSimilarity2011} is used to quantify the quality of the super-resolved image.
FSIM is a full reference image similarity metric that quantifies structural similarity, rather than individual cell values.
%TODO complain about SSIM Structural Similarity (SIM) \parencite{wangImageQualityAssessment2004}

\section{Results}
\label{sec:2results}
Model performance is quantified using FSIM, and it is seen that fine-tune training increases FSIM score for the task of super-resolution for aeromagnetic grids over the baseline model.

\begin{table}[hbt]
    \includegraphics[width=\linewidth]{fig/p2/temp_tbl.png}
    \begin{tabular}{ll}
                   & Reserved Synthetic Dataset & Reserved Aeromagnetic Dataset \\
                   & Mean / Standard Dev.       & Mean / Standard Dev.          \\
        \hline{}
        Baseline   & 0                          & 0                             \\
        Fine-Tuned & 0                          & 0                             \\
    \end{tabular}
    \caption[Accuracy Metrics]{Accuracy metrics for each upsampling method. The best performing method for each metric is bolded.}
    \label{tab:3metrics}
\end{table}

Qualitative results are shown to compare the outputs from the two models, the baseline model that is trained with synthetic data and the fine-tuned model trained with open access survey data.

@@

\begin{figure}[hbt]
    \includegraphics[width=0.9\linewidth]{fig/p2/srcomp_23.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata23}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=0.9\linewidth]{fig/p2/srcomp_14.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata14}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=0.9\linewidth]{fig/p2/srcomp_17.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata17}
\end{figure}

\begin{figure}[hbt]
    \includegraphics[width=0.9\linewidth]{fig/p2/srcomp_18.pdf}
    \caption[Super-resolution geophysics grid results]{Super-resolution predictions on open access data, prepared using the proposed method.
        Explanation in-text.
        Grid A is the super-resolution prediction using the baseline model, while grid B is the SR prediction using the fine-tuned model.}
    \label{fig:srdata18}
\end{figure}

### 2023-10-10 paper3.tex
Spell check and adjust chapter structure
Make compilable
Cleverref and subfiles seem to interact poorly, and I can't cref labels in external files in the way I had set up.

I've moved to using  \include.
Update per DW feedback

\documentclass[manuscript.tex]{subfiles}
\begin{document}
% \setcounter{chapter}{1}
\title{Implicit Neural Representation for Potential Field Geophysics}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[2]{Naveed Akhtar}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
\maketitle{}

@@

\begin{figure}[hbt]
    \centering{}
    \includegraphics[width=1.0\linewidth]{fig/p3/P864_grid_comparison_35m.pdf}
    \caption[Grid prediction outside the nominal altitude]{A constant level grid queried from the INR at the height of \qty{35}{\m} (middle).}
    % AWAGS is 80 m, but we see nothing useful at 80m
    \label{fig:grid35}
\end{figure}

@@

Inspired by the approach of \textcite{raissiPhysicsinformedNeuralNetworks2019}, a geophysics informed network could include a Laplace criterion,
\begin{equation}
    \label{eqn:cri_laplace}
    \mathcal{L}_{PDE} = \frac{1}{S_i}\sum_{i=1}^{S_i} \left(\frac{\partial{}^2f}{\partial{}x^2} + \frac{\partial{}^2f}{\partial{}y^2} + \frac{\partial{}^2f}{\partial{}z^2}\right)^2,
\end{equation}
where \(S_i\) indicates the \(i^{th}\) sample point of the field survey \(\phi(x, y, z)\).
This criterion penalises outputs where Laplace's equation is not satisfied because the sum of partial second derivatives is not close to zero.

>>>>

% % \documentclass[manuscript.tex]{subfiles}
% \documentclass[12pt,a4paper]{report} %,openright,twoside
% \usepackage{thesisstyle}
% \addbibresource{bib/PhD.bib}
% % \setcounter{chapter}{3}
% \begin{document}

\title{Implicit Neural Representation for Potential Field Geophysics}
\author[1*]{Luke Thomas Smith}
\author[1]{Tom Horrocks}
\author[2]{Naveed Akhtar}
\author[1]{Eun-Jung Holden}
\author[1]{Daniel Wedge}
\affil{Centre for Data-Driven Geoscience, School of Earth Sciences, The University of Western Australia (M006), 35 Stirling Highway, 6009 Perth, Australia}
\affil{Department of Computer Science and Software Engineering, The University of Western Australia (M002), 35 Stirling Highway, 6009 Perth, Australia}
\date{\today}
% \maketitle{}

@@

\begin{figure}[hbt]
    \centering{}
    \includegraphics[width=1.0\linewidth]{fig/p3/35_level_grid.pdf}
    \caption[Grid prediction outside the nominal altitude]{A constant level grid queried from the INR at the height of \qty{35}{\m}.}
    % AWAGS is 80 m, but we see nothing useful at 80m
    \label{fig:grid35}
\end{figure}

@@

Inspired by the approach of \textcite{raissiPhysicsinformedNeuralNetworks2019}, a geophysics-informed network could include a Laplace criterion,
\begin{equation}
    \label{eqn:cri_laplace}
    \mathcal{L}_{PDE} = \frac{1}{S_i}\sum_{i=1}^{S_i} \left(\frac{\partial{}^2f}{\partial{}x^2} + \frac{\partial{}^2f}{\partial{}y^2} + \frac{\partial{}^2f}{\partial{}z^2}\right)^2,
\end{equation}
where \(S_i\) indicates the \(i^{th}\) sample point of the field survey \(\phi(x, y, z)\).
This criterion penalises outputs where Laplace's equation is not satisfied because the sum of partial second derivatives is not close to zero.
